The \dealii{} Library, Version 9.3

### 2021-02-12 9.3/paper.tex
Add discussion related simplex, multigrid, MatrixFree, FEPointValues

@@

\hypersetup{
  pdfauthor={
     Daniel Arndt,
     Wolfgang Bangerth,
     Bruno Blais,
     Thomas C. Clevenger,
    Marc Fehling,
    Alexander V. Grayver,
    Timo Heister,
    Luca Heltai,
    Martin Kronbichler,
    Matthias Maier,
    Peter Munch,
    Jean-Paul Pelteret,
    Reza Rastak,
    Ignacio Thomas,
    Bruno Turcksin,
    Zhuoran Wang,
    David Wells
  },
  pdftitle={The deal.II Library, Version 9.2, 2020},
}

\title{The \dealii{} Library, Version 9.2}

@@

The major changes of this release are:
%
\begin{itemize}
  \item A new, parallel, fully distributed triangulation class (see
        Section~\ref{subsec:pft});
  \item Substantially improved performance for very large computations
    on tens or hundreds of thousands of processor cores and up to
    trillions of unknowns (see
        Section~\ref{subsec:performance});
  \item Better support for parallel $hp$-adaptive algorithms (see
        Section~\ref{subsec:hp});
  \item Support for particle-based methods (see
        Section~\ref{subsec:particles});
  \item Improved performance of the symbolic differentiation framework (see
        Section~\ref{subsec:symbdiff});
  \item Advances in SIMD capabilities and the matrix-free infrastructure (see
        Section~\ref{subsec:mf});
  \item Advances in GPU support (see
        Section~\ref{subsec:gpu});
  \item Better use of modern C++ language features (see
        Section~\ref{subsec:cxx});
  \item Seven new tutorial programs (see
        Section~\ref{subsec:steps}).
\end{itemize}
%
These major changes are discussed in detail in Section~\ref{sec:major}. There
are a number of other noteworthy changes in the current \dealii{} release
that we briefly outline in the remainder of this section:
%
\begin{itemize}
  \item \dealii{} had decent support for solving complex-valued problems
        for a while already
        (e.g., ones in quantum mechanics -- like the equation used in the
        \texttt{step-58} tutorial program covered below -- or for
        time-harmonic problems). However, there were two
        areas in which support was missing. First, the UMFPACK direct solver
        packaged with \dealii{} did not support solving complex-valued
        linear problems. This has now been addressed: UMFPACK actually can
        solve such systems, we just needed to write the appropriate
        interfaces. Second, the \texttt{DataOut} class that is responsible
        for converting nodal data into information that can then be written
        into files for visualization did not know how to deal with vector-
        and tensor-valued fields whose components are complex numbers. An
        example for this is to solve the time-harmonic version of the
        Maxwell equations that has the electric and magnetic fields as
        solution. This, too, has been addressed in this release.
  \item The new \texttt{DiscreteTime} class provides a more
        consistent, more readable, and less error-prone approach to control
        time-stepping algorithms within time-dependent simulations.
        While providing a rich read-only interface, the non-const interface
        of this class is designed to be minimal
        to enforce a number of important programming invariants, reducing
        the possibility of mistakes in the user code.
        For instance, \texttt{DiscreteTime} ensures that the final time step ends
        precisely on a predefined end time, automatically
        lengthening or shortening the final time step.
  \item A key component of \dealii{} are the \texttt{FEValues} and
        \texttt{FEFaceValues} classes that evaluate finite element functions
        at quadrature points located on cells and faces of a cell,
        respectively~\cite{BangerthHartmannKanschat2007}. This release now
        contains a class \texttt{FEInterfaceValues} that considers the
        restriction of the shape functions from \textit{both} sides of a face and
        allows evaluating jumps and averages of shape functions along this
        face. These are common components of the bilinear forms of
        discontinuous Galerkin schemes (as well as schemes for fourth-order
        equations, see the discussion of \texttt{step-47} below) and greatly
        simplify the implementation of these methods.
  \item Previously, the \texttt{parallel::distributed::ContinuousQuadratureDataTransfer}
        class, which transfers local quadrature point data onto the
        children of newly refined cells, did not allow different cells
        to store quadrature data of different lengths. This release lifts
        this restriction, which enables efficient quadrature data storage and data
        transfer within a triangulation containing multiple material models, each with
        their own number of local state variables and history variables.
        As an example, in a solid mechanics simulation, we can assign a hyper-elastic
        material model to a region of the mesh with no associated
        history variables, while in another part of the triangulation we incorporate an
        elasto-plastic constitutive model, requiring the storage of local plasticity
        data at quadrature points.
        During each refinement of the mesh, we can then use
        \texttt{ContinuousQuadratureDataTransfer} to transfer and interpolate
        plasticity data from parent cells to their children, ignoring the cells
        with the hyper-elastic constitutive model.
\end{itemize}
%
The changelog lists more than 240 other
features and bugfixes.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{A new fully distributed triangulation class}
\label{subsec:pft}

Previously, all triangulation classes of \dealii{} had in common that the coarse grid is replicated by
all processes in a parallel environment, and the actual mesh used for computations is constructed by repeated
refinement. However, this has its limitations in many
applications where the mesh comes
from an external mesh generator in the form of a file that frequently
already contains millions
or tens of millions of cells. For such configurations, applications might exhaust
available memory already while reading the mesh on each MPI process.

The new \texttt{parallel::fullydistributed::Triangulation} class targets this issue
by distributing also the coarse grid. Such
a triangulation can be created by providing to each process a \texttt{Triangulation\-De\-scrip\-tion::Description} struct, containing
(i) the relevant data to construct the local part of the coarse grid, (ii) the
translation of the local coarse-cell IDs to globally unique IDs, (iii) the hierarchy
of mesh refinement steps, and (iv) the owner of the cells on the active mesh level as well
as on the multigrid levels. For the current release, triangulations set up
this way cannot be adaptively refined after construction, though we plan to
improve this for the next release.

%The \texttt{TriangulationDescription::Description} struct can be filled manually or
%by the utility functions from the \texttt{TriangulationDescription::Utilities}
%namespace. The function \texttt{create\_\allowbreak description\_\allowbreak
%from\_ \allowbreak triangulation()} can convert a base triangulation (partitioned
%serial \texttt{Tri\-angulation} and \texttt{parallel::distributed::Triangulation})
%to such a struct. The advantage of this approach is that all known utility
%functions from the namespaces \texttt{GridIn} and \texttt{GridTools} can be used
%on the base triangulations before converting them to the structs. Since this
%function suffers from the same main memory problems as described above, we also
%provide the function \texttt{create\_description\_from\_triangulation\_in\_groups()},
%which creates the structs only on the master process in a process group. These
%structs are filled one by one and are sent to the relevant processes once they are
%ready. A sensible process group size might contain all processes of one compute node.

The new fully distributed triangulation class supports 1D, 2D, and 3D meshes
including geometric multigrid hierarchies, periodic boundary conditions, and
hanging nodes.

%We intend to extend the usability of the new triangulation class in regard of
%different aspects, e.g., I/O. In addition, we would like to enable adaptive mesh
%refinement, a feature of the other triangulation classes, which is very much
%appreciated by many users. For repartitioning, we plan to use an oracle approach
%known from \texttt{parallel::distributed::Triangulation}. Here, we would like to
%rely on a user-provided partitioner, which might also be a graph partitioner.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Improved large-scale performance}
\label{subsec:performance}

Large-scale simulations with up to 304,128 cores have revealed bottlenecks in release
9.1 during initialization of a number of distributed data structures, due to the usage of expensive collective operations
like \texttt{MPI\_Allgather()} and \texttt{MPI\_\allowbreak
  Alltoall()}. Typical examples are the
pre-computation of the indices of those vector entries (or
other linear index ranges) owned by
each process, which were previously stored in  an array on every process.
This information is needed to set up
the  \texttt{Utilities::MPI::Par\-ti\-ti\-oner} class.
In release 9.2, we have replaced these functions in favor of
consensus algorithms~\cite{hoefler2010scalable}, which can be
found in the namespace \texttt{Utilities::\allowbreak MPI::\allowbreak ConsensusAlgorithms} (short: \texttt{CA}).
Now, only the locally relevant information about the index ranges is
(re)computed when needed, which, for more than 100 MPI processes, uses
point-to-point communications and a single \texttt{MPI\_IBarrier()}.

%Consensus algorithms are algorithms dedicated to efficient dynamic-sparse
%communication patterns. In this context, the term ``dynamic-sparse'' means
%that by the time this function is called, the other processes do not know
%yet that they have to answer requests and
%each process only has to communicate with a small subset of processes of the
%MPI communicator. We provide two flavors of the consensus algorithm: the two-step
%approach \texttt{ConsensusAlgorithms::PEX} and the \texttt{ConsensusAlgorithms::NBX},
%which uses only point-to-point communications and a single \texttt{MPI\_IBarrier()}.
%The class \texttt{ConsensusAlgorithms::Selector} selects one of the two previous
%algorithms, depending on the number of processes.

Users can apply the new algorithms for their own dynamic-sparse problems by
providing a list of target
processes and pack/unpack routines either by implementing the interface
\texttt{CA::\allowbreak Process} or by providing \texttt{std::function}
objects to \texttt{CA::AnonymousProcess}.

By replacing the collective communications during set up and removing the arrays
that contain information for each process (enabled by the application of consensus
algorithms and other modifications---a full list of modifications leading to this
improvement can be found online), we were able to significantly
improve the set up
time for large-scale simulations and to solve a Poisson problem with multigrid
with \num{2.1e12} unknowns on the SuperMUC-NG supercomputer with 304,152
cores~\cite{dealII2020design,Arndt2020sppexa}.
Figure~\ref{fig:init_costs} compares timings of simulations of various problem
sizes (including set up) on 49,152 MPI ranks using a matrix-free
solver~\cite{Arndt2020sppexa,KronbichlerKormann2019,KronbichlerWall2018}; this solver uses discontinuous elements of
degree $5$ in a geometric multigrid (GMG) scheme. The comparison between the previous release 9.1
and the current release 9.2 shows that while the scaling for the V-cycle had been
very good before, many initialization routines have been considerably
improved, especially the enumeration of unknowns on the multigrid levels and
the setup of the multigrid transfer.

\pgfplotstableread{
  dofs  grid  distdofs  mfsetup  mgtransfer  smoother  vcycle
  14155776  0.347457  0.02424084  0.03346427  0.0105426  0.0108817  0.0035033
  28311552  0.35134  0.0215024  0.0340963  0.0095201  0.010616  0.0035887
  56623104  0.41577  0.0243593  0.0432297  0.011557  0.011667  0.0041244
  113246208  0.42172  0.0323919  0.0427806  0.012983  0.011834  0.0043853
  226492416  0.43565  0.0404079  0.0421606  0.012867  0.017305  0.0056340
  452984832  0.5055  0.0500101  0.0562977  0.017739  0.022763  0.0061992
  905969664  0.54552  0.065865  0.0641994  0.019445  0.037431  0.0087932
  1811939328  0.5672  0.104909  0.080807  0.025395  0.063548  0.0133252
  3623878656  0.65541  0.165071  0.129717  0.042294  0.12644  0.0218619
  7247757312  0.74521  0.251417  0.221072  0.073314  0.21807  0.0339264
  14495514624  0.84789  0.403463  0.36894  0.11425  0.4091  0.0669798
  28991029248  0.91983  0.52435  0.5363  0.14189  0.79178  0.1322448
  57982058496  1.0339  0.87448  0.93134  0.23863  1.5896  0.2539445
  115964116992  1.2667  1.46016  1.70806  0.44101  3.1409  0.4956625
  231928233984  1.7597  2.58121  3.3741  0.77624  6.2628  0.9857200
}\tableinit

\pgfplotstableread{
  dofs  grid  distdofs  mfsetup  mgtransfer  smoother  vcycle
  14155776  2.0148  0.9086  0.433886  1.3099  0.010685  0.0035033
  28311552  1.8746  0.91078  0.576777  1.3093  0.010752  0.0035887
  56623104  2.0924  0.98058  0.574183  1.4459  0.011831  0.0041244
  113246208  2.3788  1.22025  0.547492  1.7919  0.011979  0.0043853
  226492416  2.1254  1.50329  0.839219  1.7832  0.017819  0.0056340
  452984832  2.4325  2.10171  0.711708  2.0263  0.023393  0.0061992
  905969664  3.8911  3.58355  1.08478  2.5844  0.045697  0.0087932
  1811939328  4.1803  4.379  0.784804  1.8947  0.061552  0.0133252
  3623878656  2.6654  5.6073  0.833896  1.8912  0.1188  0.0218619
  7247757312  2.7422  7.8463  0.741733  1.8477  0.21291  0.0339264
  14495514624  2.765  9.9848  0.99218  1.8248  0.40665  0.0669798
  28991029248  3.2901  15.0526  1.5124  2.6723  0.81013  0.1322448
  57982058496  3.1927  21.0562  1.72831  2.1998  1.5905  0.2539445
  115964116992  4.1485  31.0087  2.67998  2.9091  3.2009  0.4956625
  231928233984  4.1787  47.195  4.2821  3.9558  6.3362  0.9857200
}\tableinitold

\begin{figure}
  \begin{tikzpicture}
    \begin{loglogaxis}[
        width=0.51\textwidth,
        height=0.38\textwidth,
        xlabel={\# DoFs},
        ylabel={time [s]},
        title={9.1 release},
        xtick={1.4e7,5.7e7,2.3e8,9.1e8,3.6e9,1.4e10,5.8e10,2.3e11},
        xticklabels={1.4$\times 10^7$,5.7$\times 10^7$,2.3$\times 10^8$,9.1$\times 10^8$,3.6$\times 10^9$,1.4$\times 10^{10}$,5.8$\times 10^{10}$,2.3$\times 10^{11}$},
        tick label style={font=\scriptsize},
        xticklabel style={rotate=25,anchor=north east},
        label style={font=\scriptsize},
        legend style={font=\scriptsize},
        %legend to name=legend:scalingMG,
        %legend cell align=left,
        %legend columns = 3,
        ymin=2e-3,ymax=8e1,
        grid, thick]
      \addplot table[x index = 0, y index={1}] {\tableinitold};
      \addplot[red,mark=square*] table[x index = 0, y index={2}] {\tableinitold};
      \addplot[gnuplot@green,mark=otimes] table[x index = 0, y index={3}] {\tableinitold};
      \addplot[gnuplot@lightblue,mark=triangle] table[x index = 0, y index={4}] {\tableinitold};
      \addplot[gnuplot@purple,mark=diamond] table[x index = 0, y index={5}] {\tableinitold};
      \addplot[black,mark=x] table[x index = 0, y index={6}] {\tableinitold};
    \end{loglogaxis}
  \end{tikzpicture}
  \hfill
  \begin{tikzpicture}
    \begin{loglogaxis}[
        width=0.51\textwidth,
        height=0.38\textwidth,
        xlabel={\# DoFs},
        ylabel={time [s]},
        title={9.2 release},
        xtick={1.4e7,5.7e7,2.3e8,9.1e8,3.6e9,1.4e10,5.8e10,2.3e11},
        xticklabels={1.4$\times 10^7$,5.7$\times 10^7$,2.3$\times 10^8$,9.1$\times 10^8$,3.6$\times 10^9$,1.4$\times 10^{10}$,5.8$\times 10^{10}$,2.3$\times 10^{11}$},
        tick label style={font=\scriptsize},
        xticklabel style={rotate=25,anchor=north east}, thick,
        label style={font=\scriptsize},
        legend style={font=\scriptsize},
        legend to name=legend:init,
        legend cell align=left,
        legend columns = 6,
        ymin=2e-3,ymax=8e1,
        grid]
      \addplot table[x index = 0, y index={1}] {\tableinit};
      \addlegendentry{create mesh};
      \addplot[red,mark=square*] table[x index = 0, y index={2}] {\tableinit};
      \addlegendentry{enumerate (mg) dofs};
      \addplot[gnuplot@green,mark=otimes] table[x index = 0, y index={3}] {\tableinit};
      \addlegendentry{init matrix-free};
      \addplot[gnuplot@lightblue,mark=triangle] table[x index = 0, y index={4}] {\tableinit};
      \addlegendentry{init GMG transfer};
      \addplot[gnuplot@purple,mark=diamond] table[x index = 0, y index={5}] {\tableinit};
      \addlegendentry{init smoother};
      \addplot[black,mark=x] table[x index = 0, y index={6}] {\tableinit};
      \addlegendentry{GMG V-cycle};
    \end{loglogaxis}
  \end{tikzpicture}
  \\
  \strut\hfill\pgfplotslegendfromname{legend:init}\hfill\strut
  \caption{\it Comparison of initialization costs of various data structures in the 9.1 release (left) and the new 9.2 release (right) when run on 49,152 MPI ranks.}
  \label{fig:init_costs}
\end{figure}

As part of this effort, we ran benchmarks on the TACC Frontera system, where we were
able to apply the matrix-free geometric multigrid framework to a variable viscosity
Stokes system and achieved weak and strong scaling up to 114K MPI ranks with up to
\num{2.1e11} unknowns. This is likely the largest block system currently solved with
\dealii{} and required various optimizations
and fixes on top of the ones mentioned above:
(i) Bug fixes to concurrent point to point communications.
(ii) Fixes to multigrid transfer with adaptive refinement and more than \num{4e9} unknowns.
(iii) Fixes to index sets in block indices with more than \num{4e9} unknowns.
(iv) Fixes to computations with more than \num{4e9} active cells.
(v) Implementation of IDR(s) solvers to reduce memory overhead.
For more details, see \cite{clevenger_stokes19}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Better support for parallel $hp$-adaptive algorithms}
\label{subsec:hp}

Since the previous release, \dealii{} has had support for $hp$-adaptive finite element methods
on distributed memory systems \cite{dealII91}. We implemented the bare functionality
for $hp$-adaptive methods with the objective to offer the greatest flexibility in
their application. Here, reference finite elements still had to be assigned manually
to each cell, which may not lead to an optimal mesh and is tedious.

With the current release, $hp$-adaptive finite element methods have been further
expanded: New features like decision strategies have been added and the user interface
has been overhauled, effectively making $hp$-methods more attractive to use. We introduced
many new functions that automatize the general workflow for applying $hp$-decision
strategies, which run on top of the previous low-level implementation for both serial
and parallel applications.

>>>>

\usetikzlibrary{shapes.misc}
\tikzset{cross/.style={cross out, draw=black, minimum size=2*(#1-\pgflinewidth), inner sep=0pt, outer sep=0pt},
%default radius will be 1pt. 
cross/.default={2pt}}

@@

\hypersetup{
  pdfauthor={
     Daniel Arndt,
     Wolfgang Bangerth,
     Bruno Blais,
     Thomas C. Clevenger,
    Marc Fehling,
    Alexander V. Grayver,
    Timo Heister,
    Luca Heltai,
    Martin Kronbichler,
    Matthias Maier,
    Peter Munch,
    Jean-Paul Pelteret,
    Reza Rastak,
    Ignacio Thomas,
    Bruno Turcksin,
    Zhuoran Wang,
    David Wells
  },
  pdftitle={The deal.II Library, Version 9.3, 2021},
}

\title{The \dealii{} Library, Version 9.3}

@@

The major changes of this release are:
%
\begin{itemize}
  \item A new, parallel, fully distributed triangulation class (see
        Section~\ref{subsec:pft});
  \item Substantially improved performance for very large computations
    on tens or hundreds of thousands of processor cores and up to
    trillions of unknowns (see
        Section~\ref{subsec:performance});
  \item Better support for parallel $hp$-adaptive algorithms (see
        Section~\ref{subsec:hp});
  \item Support for particle-based methods (see
        Section~\ref{subsec:particles});
  \item Improved performance of the symbolic differentiation framework (see
        Section~\ref{subsec:symbdiff});
  \item Advances in SIMD capabilities and the matrix-free infrastructure (see
        Section~\ref{subsec:mf});
  \item Advances in GPU support (see
        Section~\ref{subsec:gpu});
  \item Better use of modern C++ language features (see
        Section~\ref{subsec:cxx});
  \item Seven new tutorial programs (see
        Section~\ref{subsec:steps}).
\end{itemize}
%
These major changes are discussed in detail in Section~\ref{sec:major}. There
are a number of other noteworthy changes in the current \dealii{} release
that we briefly outline in the remainder of this section:
%
\begin{itemize}
  \item globally unique cell IDs
\end{itemize}
%
The changelog lists more than 240 other
features and bugfixes.

@@

\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Simplex and mixed mesh support}
\label{subsec:simplex}

\begin{figure}[!h]

\centering

\begin{tikzpicture}[scale=1.7]

\coordinate (0) at (0,0);
\coordinate (1) at (1,0);
\coordinate (2) at (0,1);

\coordinate (3) at (0,0.5);
\coordinate (4) at (0.5,0.5);
\coordinate (5) at (0.5,0);

\foreach \i in {0,1, 2}
    \fill (\i) circle (1.1pt) node [below] {};

\foreach \i in {3,4,5}
    \draw (\i) node[cross, draw=red, rotate=-20] {};

\draw (0) --(1) -- (2) -- (0);

\end{tikzpicture}
\begin{tikzpicture}[scale=1.7]

\coordinate (0) at (0,0);
\coordinate (1) at (1,0);
\coordinate (2) at (1,1);
\coordinate (3) at (0,1);

\coordinate (4) at (0.5,0,0);
\coordinate (5) at (1.0,0.5);
\coordinate (6) at (0.0,0.5);
\coordinate (7) at (0.5,1.0);
\coordinate (8) at (0.5,0.5);

\foreach \i in {0,1, 2, 3}
    \fill (\i) circle (1.1pt) node [below] {};

\foreach \i in {4,5,6,7,8}
    \draw (\i) node[cross, draw=red, rotate=-20] {};

\draw (0) --(1) -- (2) -- (3) -- (0);

\end{tikzpicture}
\qquad\qquad
\begin{tikzpicture}[scale=1.3]

\coordinate (0) at (0,0,0);
\coordinate (1) at (1,0,0);
\coordinate (2) at (0,1,0);
\coordinate (3) at (0,0,1);

\coordinate (4) at (0.5,0,0);
\coordinate (5) at (0,0.5,0);
\coordinate (6) at (0,0,0.5);
\coordinate (7) at (0.5,0.5,0.0);
\coordinate (8) at (0.5,0.0,0.5);
\coordinate (9) at (0.0,0.5,0.5);

\foreach \i in {0,1,2,3}
    \fill (\i) circle (1.5pt) node [below] {};

\foreach \i in {4,5,6,7,8,9}
    \draw (\i) node[cross, draw=red, rotate=-20] {};

\draw (0) -- (1) -- (2) -- (0);
\draw (0) -- (1) -- (3) -- (0);
\draw (1) -- (2) -- (3) -- (1);

\end{tikzpicture}
\begin{tikzpicture}[scale=1.5]

\coordinate (0) at (0.0, 0.0, 0.0);
\coordinate (1) at (0.0, 0.0, 1.0);
\coordinate (2) at (1.0, 0.0, 0.0);
\coordinate (3) at (1.0, 0.0, 1.0);
\coordinate (4) at (0.5, 1.0, 0.5);

\coordinate  (5) at (0.5, 0.0, 0.0);
\coordinate  (6) at (0.0, 0.0, 0.5);
\coordinate  (7) at (0.5, 0.0, 1.0);
\coordinate  (8) at (1.0, 0.0, 0.5);

\coordinate  (9) at (0.25,0.5,0.25);
\coordinate (10) at (0.25,0.5,0.75);
\coordinate (11) at (0.75,0.5,0.25);
\coordinate (12) at (0.75,0.5,0.75);

\coordinate (13) at (0.5, 0.0, 0.5);

\foreach \i in {0,1,2,3,4}
    \fill (\i) circle (1.3pt) node [below] {};

%\foreach \i in {5, 6, 7, 8, 9, 10, 11, 12, 13}
%    \draw (\i) node[cross] {};

\draw (0) -- (1) -- (3) -- (2) -- (0);
\draw (1) -- (4) -- (3);
\draw (0) -- (4) -- (2);

\end{tikzpicture}
\begin{tikzpicture}[scale=1.3]

\coordinate (0) at (0,0,1.5);
\coordinate (1) at (1,0,1.5);
\coordinate (2) at (0.0,1.0,1.5);

\coordinate (3) at (0,0,0);
\coordinate (4) at (1,0,0);
\coordinate (5) at (0.0,1.0,0);

\coordinate  (6) at (0.5, 0.0, 0.0);
\coordinate  (7) at (0.0, 0.5, 0.0);
\coordinate  (8) at (0.5, 0.5, 0.0);

\coordinate   (9) at (0.5, 0.0, 0.75);
\coordinate  (10) at (0.0, 0.5, 0.75);
\coordinate  (11) at (0.5, 0.5, 0.75);

\coordinate  (12) at (0.5, 0.0, 1.5);
\coordinate  (13) at (0.0, 0.5, 1.5);
\coordinate  (14) at (0.5, 0.5, 1.5);

\foreach \i in {0,1,2,3,4,5}
    \fill (\i) circle (1.5pt) node [below] {};

\foreach \i in {6,7,8,9,10,11,12,13,14}
    \draw (\i) node[cross, draw=red, rotate=-20] {};

\draw (0) -- (1) -- (2) -- (0);
\draw (3) -- (4) -- (5) -- (3);
\draw (0) -- (3);
\draw (1) -- (4);
\draw (2) -- (5);

\end{tikzpicture}
\begin{tikzpicture}[scale=1.5]

\coordinate (0) at (0,0,1);
\coordinate (1) at (1,0,1);
\coordinate (2) at (0,0,0);
\coordinate (3) at (1,0,0);

\coordinate (4) at (0,1,1);
\coordinate (5) at (1,1,1);
\coordinate (6) at (0,1,0);
\coordinate (7) at (1,1,0);

\coordinate  (8) at (0,0,0.5);
\coordinate  (9) at (1,0,0.5);
\coordinate (10) at (0.5,0,1);
\coordinate (11) at (0.5,0,0);
\coordinate (12) at (0.5,0,0.5);

\coordinate (13) at (0,1,0.5);
\coordinate (14) at (1,1,0.5);
\coordinate (15) at (0.5,1,1);
\coordinate (16) at (0.5,1,0);
\coordinate (17) at (0.5,1,0.5);

\coordinate (18) at (0,0.5,1);
\coordinate (19) at (1,0.5,1);
\coordinate (20) at (0,0.5,0);
\coordinate (21) at (1,0.5,0);
\coordinate (22) at (0,0.5,0.5);
\coordinate (23) at (1,0.5,0.5);
\coordinate (24) at (0.5,0.5,1);
\coordinate (25) at (0.5,0.5,0);
\coordinate (26) at (0.5,0.5,0.5);

\foreach \i in {0,1,2,3,4,5,6,7}
    \fill (\i) circle (1.3pt) node [below] {};

\foreach \i in {8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24,25,26}
    \draw (\i) node[cross, draw=red, rotate=-20] {};

\draw (0) -- (1) -- (3) -- (2) -- (0);
\draw (4) -- (5) -- (7) -- (6) -- (4);
\draw (0) -- (4);
\draw (1) -- (5);
\draw (2) -- (6);
\draw (3) -- (7);

\end{tikzpicture}

\caption{All reference cells in 2D (triangle, quadrilateral) and 3D (tetrahedron, pyramid, wedge, hexahedron) with the support points indicated both for linear ($\bullet$) and quadratic ($\times$) shape functions.}

\end{figure}

\begin{table}[!h]
  \caption{\it List of new scalar \texttt{FiniteElement} classes for the new
  reference-cell types. Vectorial elements can be constructed based on these classes
  via \texttt{FE\_Systems}.}\label{tab:simplex:fe}
  \centering
  \begin{tabular}{cccc}
    \toprule
    \textbf{reference cell} & \textbf{finite element} & \textbf{dimension} & \textbf{degree}\\
    \midrule
    simplex (line, triangle, tetrahedron) & FE\_P, FE\_DGP  & 1--3 & 1--2 \\
    pyramid & FE\_PyramidP, FE\_PyramidDGP  & 3 & 1 \\
    wedge & FE\_WedgeP, FE\_WedgeDGP  & 3 & 1--2 \\
    \bottomrule
  \end{tabular}
\end{table}

The current release of \dealii adds support for simplex meshes (consisting of triangles in 2D; tetrahedra in 3D) and mixed meshes (consisting of triangles and/or 
quadrilaterals in 2D; tetrahedra, pyramids, wedges, and/or hexahedra in 3D). 
Many freely available mesh-generation tools produce such kind of meshes; while users of \dealii used to have to pre-process such meshes and convert them to pure hex meshes, they
can now directly work with them.

\subsubsection{Refactoring of internal data structures}

To enable simplex and mixed mesh support, we performed a major refactoring of the internal data structures of \dealii. In particular, the \texttt{Triangulation} class and 
the \texttt{DoFHandler} class have undergone large changes.

The template parameters of the internal data structures of \texttt{Triangulation}
have been removed and the type of each cell and of each face (only in 3D) is stored. The function
\texttt{Triangulation::create\_\allowbreak triangulation()}, which converts a given list of
cells and vertices to the internal data structures, has been rewritten 
inspired by \citep{logg09} and a speed-up of up to 5 has been reached. Minor adjustments 
have been made to \texttt{parallel::shared::Triangulation} and \texttt{parallel::\allowbreak fullydistributed::\allowbreak Triangulation} such that 
the new mesh types can be processed also in parallel.

The internal data structures of \texttt{DoFHandler} used to be hard-coded for pure
hypercube meshes, while the \texttt{hp::DoFHandler} used to be built around
CRS-like data structures. Due to the need of CRS data structures in
\texttt{DoFHandler} in the context of more general meshes, we have merged
\texttt{hp::DoFHandler} into the \texttt{DoFHandler}. The class \texttt{hp::DoFHandler}, which is now a dummy derivation of \texttt{DoFHandler}, currently only exists for compatibility reasons. For
more details see Subsection~\ref{TODO}.

\subsubsection{Generating meshes }

The most obvious way to generate a simplex or a mixed mesh is to read the mesh from a file
generated by an external mesh generator. Currently, we support the 
file formats \texttt{VTK}, \texttt{MSH}, \texttt{EXODUS II}.

Alternatively, one can create a pure hypercube mesh with the known functions
in the \texttt{GridGenerator} namespace and convert the obtained mesh to a
pure simplex mesh with the function 
\texttt{convert\_\allowbreak hypercube\_\allowbreak to\_\allowbreak simplex\_\allowbreak mesh()} from the \texttt{GridGenerator} namespace.

\subsubsection{Simplex mesh}

After having created a triangulation, one can proceed as usual in the case of
a pure simplex mesh by selecting an appropriate finite element, mapping, and
quadrature class:

\begin{c++}
Simplex::FE_P<dim, spacedim> fe(degree);
MappingFE<dim, spacedim> mapping(Simplex::FE_P<dim, spacedim>(1));
Simplex::QGauss<dim> quad(degree + 1);

DoFHandler<dim> dof_handler(tria);
dof_handler.distribute_dofs(fe);

FEValues<dim, spacedim> fe_values(mapping, fe, quad, flags);
\end{c++}
The  list of currently supported finite-element classes is provided in Table~\ref{tab:simplex:fe}. Currently,
only linear iso-parametric mapping (\texttt{MappingFE}) is available, which takes one of the 
finite elements mentioned before. For quadrature, the classes \texttt{QGauss}, 
\texttt{QGaussPyamid}, and \texttt{QGaussWedge} from the \texttt{simplex} namespace are
available.

\subsubsection{Mixed mesh}
For mixed meshes, concepts known from the hp context have been applied: 
 different finite-element classes are assigned to different cell types. In the 
case of a 2D mixed mesh, which can only consists of triangles and
quadrilaterals, the finite element defined on a triangle (e.g., \texttt{Simplex::FE\_P})
and on a quadrilateral (e.g., \texttt{FE\_Q}) can be collected in a \texttt{hp::FECollection}:

\begin{c++}  
hp::FECollection<dim, spacedim> fe
  {Simplex::FE_P<dim, spacedim>(degree), FE_Q<dim, spacedim>(degree)};
\end{c++}

Similarly, \texttt{hp::QCollection} and \texttt{hp::MappingCollection} can be used
to construct appropriate collections. Furthermore, the right active finite-element index, which points to the right 
finite element of that cell, has to be assigned to each cell:

\begin{c++}  
DoFHandler<dim> dof_handler(tria);

for (const auto &cell : dof_handler.active_cell_iterators())
  switch(cell->reference_cell_type())
    {
      case ReferenceCell::Type::Tri:  cell->set_active_fe_index(0); break;
      case ReferenceCell::Type::Quad: cell->set_active_fe_index(1); break;
      // 3D (Tet, Pyramid, Wedge, Hex) not shown
      default: Assert(false, ExcNotImplemented());
    }

dof_handler.distribute_dofs(fe);
\end{c++}

\subsubsection{Practical implications}

The introduction of simplex and mixed meshes leads to some implications
for the user if these features should be used. For instance, each cell might have a different type with 
different number of vertices, lines, and faces so that these quantities cannot be 
compile-time constants anymore. This information used to be queried from
the \texttt{GeometryInfo} class. To avoid using this class, we have extended
relevant classes, e.g., \texttt{TriaAccessor} or \texttt{TriaCellAccessor},
with useful new functions like \texttt{n\_vertices()}, \texttt{n\_lines()}, or
\texttt{n\_faces()} so that users can simply write:
\begin{c++}
for(const auto & cell : tria.active_cell_iterators())
  for(unsigned int f = 0; f < cell->n_faces(); ++f)
    (void) cell->face(f);
\end{c++}
Alternatively, one can use an iterator-based approach to loop over all faces
of a cell, introduced in the last release. The relevant functions have been adjusted to be able to deal with
various number of faces. 
\begin{c++}
for(const auto & cell : tria.active_cell_iterators())
  for(const auto & face : cell->face_iterators())
    (void) face;
\end{c++}
Furthermore, the number of degrees of freedom might differ between cells so that
local data storage units needed for assembly might need to be resized for each 
cell as it is already good practice in the hp context:
\begin{c++}
std::vector<double> local_rhs;
for(const auto & cell : tria.active_cell_iterators())
 {
   hp_fe_values.reinit(cell);
   local_rhs.resize(hp_fe_values.get_present_fe_values().n_dofs_per_cell());
 }
\end{c++}

What is true for cells is also true for faces in 3D: faces can be either triangles
or quadrilaterals. This is the case even if no mixed mesh is used. As a consequence, some functions, e.g. \texttt{FiniteElementData::n\_dofs\_per\_face()}, have been extended with a new optional argument for the face number.

We tried---where possible---to provide utility functions so that users do not need to access a \texttt{GeometryInfo}-like data structure. However, these functions internally rely on the \texttt{ReferenceCell::\allowbreak internal::\allowbreak Info} classes, which are similarly structured as \texttt{GeometryInfo} and even use it for hypercube cells. Users can query the reference-cell type of geometric entity via:
\begin{c++}
const auto reference_cell_type = cell->reference_cell_type();
\end{c++}
and can access the relevant struct describing the current reference cell or its i-th face via:
\begin{c++}
ReferenceCell::internal::Info::get_cell(reference_cell_type);
ReferenceCell::internal::Info::get_face(reference_cell_type, i);
\end{c++}

Furthermore, it is crucial to use the right FE--mapping--quadrature family 
for the right cell type. In particular, this means that users cannot 
rely on default parameters regarding mapping and/or quadrature in the 
context of simplex meshes since these are defined for hypercube cells.
We recommend to choose the needed mapping,
FE, and quadrature rules at a single central place and pass these throughout the program
in the form of ``scratch data''.

\subsubsection{Matrix-free support}
